---
title: "01_DataPreProcessing"
author: "Xiaoran Sun"
date: "2019/2/11"
output: html_document
---
## Data preprocessing and preparation.
The Steps include:
1. Loading packages used, setting work directory, and loading raw data in your environment.
2. Selecting and creating family-related variables from Wave I, as the features/predictors that will be used in next-step analysis with machine learning.
3. Selecting and creating the outcome variables (i.e., college enrollment and college completion).
4. Handling missing data and sample weighting.
5. Preliminary analysis for descriptives (weighted).


## 1. Loading packages, setting work directory, and loading raw data
```{r}
#install required packages if not in the environment
list.of.packages <- c("foreign", "SDMTools", "weights")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

#loading required packages
library(foreign)
library(SDMTools)
library(weights)

#setting work directory
##setwd("~/Desktop/paper/SunProjects/Add Health") #Xiaoran's work directory; please set your own.

#loading raw data
##reading in Wave 1 In-Home Interview Data
load("Wave1InHome.rda")  #R version
wave1<-da21600.0001
spss1<-data.frame(read.spss("21600-0001-Data.sav")) #SPSS version
dat1<-wave1[c("AID")] #create this data called 'dat1', where all the cleaned variables will be stored
                      #6504 observations

#loading Wave 4 data
load("Wave4InHome.rda")
wave4<-da21600.0022

#loading Wave 4 weights
load("Wave4Weights.rda") #Use Wave IV weights instead of Wave I weights. See example w/ Wave III data: Perreira, Krista M.; Harris, Kathleen Mullan; & Lee, Dohoon. (2006).
Wave4Weights<-da21600.0031
```

## 2. Selecting and creating family-related variables from Wave I
The variables (in total, 55) include:
* Family and adolescent demographic characteristics (9 variables)
* Family socioeconomic characteristics (8 variables);
* Family and parent-adolescent relationship characteristics (10 variables);
* Parental involvement with education (7 variables);
* Family sociocultural characteristics (10 variables);
* Family health resources and behaviors (11 variables).

In the meantime, mean imputation for skip pattern missingness in these variables.

### Family and adolescent demographic characteristics
Resident mother/father presence
```{r}
Mother<-rep(1, 6504)
Mother[which(as.numeric(spss1$H1RM1)==14)]<-0
Father<-rep(1, 6504)
Father[which(as.numeric(spss1$H1RF1)==14)]<-0

dat1$Mother<-Mother
dat1$Father<-Father
```
Resident with two biological parents
```{r}
levels(spss1$H1NM1)
BioMom<-rep(NA, 6504)
BioMom[which(as.numeric(spss1$H1NM1)==4)]<-1
BioMom[which(as.numeric(spss1$H1NM1)<3)]<-0

levels(spss1$H1NF1)
BioDad<-rep(NA, 6504)
BioDad[which(as.numeric(spss1$H1NF1)==4)]<-1
BioDad[which(as.numeric(spss1$H1NF1)<3)]<-0

TwoBioParent<-rep(0, 6504)
TwoBioParent[(which(BioMom+BioDad==2))]<-1
dat1$TwoBioParent<-TwoBioParent
```

Household size
```{r}
levels(spss1$H1HR2A)
which(as.numeric(spss1$H1HR2A)==4)
levels(spss1$H1HR2E)

HHsize<-rep(NA,6504)
HHsize[which(as.numeric(spss1$H1HR2G)!=4)]<-7 #lives with seven or more
HHsize[which(as.numeric(spss1$H1HR2F)!=4 & as.numeric(spss1$H1HR2G)==4)]<-6 #lives with exactly six
HHsize[which(as.numeric(spss1$H1HR2E)!=4 & as.numeric(spss1$H1HR2F)==4)]<-5 #lives with exactly five
HHsize[which(as.numeric(spss1$H1HR2D)!=4 & as.numeric(spss1$H1HR2E)==4)]<-4
HHsize[which(as.numeric(spss1$H1HR2C)!=4 & as.numeric(spss1$H1HR2D)==4)]<-3
HHsize[which(as.numeric(spss1$H1HR2B)!=4 & as.numeric(spss1$H1HR2C)==4)]<-2
HHsize[which(as.numeric(spss1$H1HR2A)!=4 & as.numeric(spss1$H1HR2B)==4)]<-1
HHsize[which(as.numeric(spss1$H1HR2A)==4)]<-0 #lives alone
summary(as.factor(HHsize))

dat1$HHsize<-HHsize
```

Number of siblings
```{r}
levels(spss1$H1HR3A)
SibNum<-rep(0,6504)
SibA<-(as.numeric(spss1$H1HR3A)==5|as.numeric(spss1$H1HR3A)==8)
SibB<-(as.numeric(spss1$H1HR3B)==5|as.numeric(spss1$H1HR3B)==8)
SibC<-(as.numeric(spss1$H1HR3C)==5|as.numeric(spss1$H1HR3C)==8)
SibD<-(as.numeric(spss1$H1HR3D)==5|as.numeric(spss1$H1HR3D)==8)
SibE<-(as.numeric(spss1$H1HR3E)==5|as.numeric(spss1$H1HR3E)==8)
SibF<-(as.numeric(spss1$H1HR3F)==5|as.numeric(spss1$H1HR3F)==8)
SibG<-(as.numeric(spss1$H1HR3G)==5|as.numeric(spss1$H1HR3G)==8)
SibH<-(as.numeric(spss1$H1HR3H)==5|as.numeric(spss1$H1HR3H)==8)
SibI<-(as.numeric(spss1$H1HR3I)==5|as.numeric(spss1$H1HR3I)==8)
SibJ<-(as.numeric(spss1$H1HR3J)==5|as.numeric(spss1$H1HR3J)==8)
SibK<-(as.numeric(spss1$H1HR3K)==5|as.numeric(spss1$H1HR3K)==8)
SibL<-(as.numeric(spss1$H1HR3L)==5|as.numeric(spss1$H1HR3L)==8)
SibM<-(as.numeric(spss1$H1HR3M)==5|as.numeric(spss1$H1HR3M)==8)
SibN<-(as.numeric(spss1$H1HR3N)==5|as.numeric(spss1$H1HR3N)==8)
SibO<-(as.numeric(spss1$H1HR3O)==5|as.numeric(spss1$H1HR3O)==8)
SibP<-(as.numeric(spss1$H1HR3P)==5|as.numeric(spss1$H1HR3P)==8)
SibNum<-SibA+SibB+SibC+SibD+SibE+SibF+SibG+SibH+SibI+SibJ+SibK+SibL+SibM+SibN+SibO+SibP
summary(SibNum)
cor(SibNum, HHsize)
head(as.numeric(spss1$H1HR3A))
dat1$SibNum<-SibNum
```

parent age
```{r}
head(wave1$PA2)  #responded by parents
dat1$PAage<-wave1$PA2
```

Birth Order
```{r}
BirthOrder<-rep(NA, 6504)
levels(spss1$H1HR14)
BirthOrder[which(as.numeric(spss1$H1HR14)==1)]<-1 #only child
levels(spss1$H1HR15)
BirthOrder[which(as.numeric(spss1$H1HR15)==1)]<-1
BirthOrder[which(as.numeric(spss1$H1HR15)==2)]<-2
BirthOrder[which(as.numeric(spss1$H1HR15)==3)]<-3
BirthOrder[which(as.numeric(spss1$H1HR15)>=4 & as.numeric(spss1$H1HR15)<=14)]<-4 #4 and above
summary(as.factor(BirthOrder))
dat1$BirthOrder<-BirthOrder
```
Biological Sex, Age
```{r}
dat1$biosex<-as.numeric(substr(wave1$BIO_SEX, start=2, stop = 2))-1
wave1$YAge<-(as.numeric(wave1$IYEAR)+1993)- (as.numeric(wave1$H1GI1Y)+1973)+((as.numeric(substr(wave1$IMONTH, start=2, stop=3)))-(as.numeric(substr(wave1$H1GI1M, start=2, stop=3))))/12
dat1$YAge<-wave1$YAge
```

### Family socioeconomic characteristics
Mother/father education levels (resident parents)
* no resident mom/dad code as mean 
```{r}
#youth report
momeduc<-as.numeric(substr(wave1$H1RM1, start = 2, stop = 3)) 
momeduc[which(momeduc==10)]<-1 #never went to school
momeduc[which(momeduc==9)]<-8
momeduc[which(momeduc==11|momeduc==12)]<-NA
mean(momeduc, na.rm = TRUE)
length(spss1[as.numeric(spss1$H1RM1)==14,]$AID) #legitimate skip
momeduc[which(as.numeric(spss1$H1RM1)==14)]<- 5.435109 #no resident mom
summary(as.factor(momeduc))

dadeduc<-as.numeric(substr(wave1$H1RF1, start = 2, stop = 3)) 
dadeduc[which(dadeduc==10)]<-1 #never went to school
dadeduc[which(dadeduc==9)]<-8
dadeduc[which(dadeduc==11|dadeduc==12)]<-NA
mean(dadeduc, na.rm = TRUE)
length(spss1[as.numeric(spss1$H1RF1)==14,]$AID) #legitimate skip
dadeduc[which(as.numeric(spss1$H1RF1)==14)]<- 5.523199 #no resident dad
summary(as.factor(dadeduc))

dat1$momeduc<-momeduc
dat1$dadeduc<-dadeduc
```

Mother/Father occupational prestige
* recode into five ordered categories (Feliciano & Lanuza, 2017):
  professionals, office/sales workers, blue-collar workers, crafts/military/farm/other, and unemployed (5 to 1)
* Note: No mom/dad code as mean
```{r}
#youth report
levels(wave1$H1RM4)
Ymomjob<-as.numeric(substr(wave1$H1RM4, start = 2, stop = 3))
Ymomjob_re<-rep(NA, 6504)
Ymomjob_re[which(Ymomjob<=4)]<-5  
Ymomjob_re[which(Ymomjob==5|Ymomjob==6)]<-4
Ymomjob_re[which(Ymomjob==7|Ymomjob==9|Ymomjob==10|Ymomjob==11|Ymomjob==12)]<-3
Ymomjob_re[which(Ymomjob==8|Ymomjob==13|Ymomjob==14|Ymomjob==15)]<-2
Ymomjob_re[which(Ymomjob==16)]<-1
mean(Ymomjob_re, na.rm = TRUE)
Ymomjob_re[which(as.numeric(spss1$H1RM4)==18)]<-3.350598
summary(as.factor(Ymomjob_re))

Ydadjob<-as.numeric(substr(wave1$H1RF4, start = 2, stop = 3))
Ydadjob_re<-rep(NA, 6504)
Ydadjob_re[which(Ydadjob<=4)]<-5  
Ydadjob_re[which(Ydadjob==5|Ydadjob==6)]<-4
Ydadjob_re[which(Ydadjob==7|Ydadjob==9|Ydadjob==10|Ydadjob==11|Ydadjob==12)]<-3
Ydadjob_re[which(Ydadjob==8|Ydadjob==13|Ydadjob==14|Ydadjob==15)]<-2
Ydadjob_re[which(Ydadjob==16)]<-1
mean(Ydadjob_re, na.rm = TRUE)
Ydadjob_re[which(as.numeric(spss1$H1RF4)==18)]<-3.347201
summary(as.factor(Ydadjob_re))

dat1$momjob<-Ymomjob_re
dat1$dadjob<-Ydadjob_re
```

Family income (in thousand, log)
```{r}
dat1$faminc<-log(wave1$PA55+1)
```

Parent receiving public assistance
```{r}
PAassistance<-as.numeric(substr(wave1$PA21, start = 2, stop = 2))
dat1$PAassistance<-PAassistance
```

Family receive welfare
```{r}
welfare1<-as.numeric(substr(wave1$PA57B, start = 2, stop = 2))
welfare2<-as.numeric(substr(wave1$PA57C, start = 2, stop = 2))
welfare3<-as.numeric(substr(wave1$PA57D, start = 2, stop = 2))
welfare<-rowSums(cbind(welfare1, welfare2, welfare3), na.rm = TRUE)
welfare[which(is.na(welfare1)&is.na(welfare2)&is.na(welfare3))]<-NA
welfare[which(welfare>=1)]<-1
dat1$welfare<-welfare
```

Parent economic hardship;
```{r}
levels(spss1$PA56)
PAeconhard<-1-as.numeric(substr(wave1$PA56, start = 2, stop = 2))
dat1$PAeconhard<-PAeconhard
```

### Family and parent-adolescent relationship characteristics
family social support (Ryabov, 2016)
```{r}
levels(wave1$H1PR3)
levels(wave1$H1PR5)
levels(wave1$H1PR7)
levels(wave1$H1PR8)
famsup1<-as.numeric(substr(wave1$H1PR3, start = 2, stop = 2))
famsup1[which(famsup1==6)]<-NA #"does not apply"
famsup2<-as.numeric(substr(wave1$H1PR5, start = 2, stop = 2))
famsup2[which(famsup2==6)]<-NA #"does not apply"
famsup3<-as.numeric(substr(wave1$H1PR7, start = 2, stop = 2))
famsup3[which(famsup3==6)]<-NA #"does not apply"
famsup4<-as.numeric(substr(wave1$H1PR8, start = 2, stop = 2))
famsup4[which(famsup4==6)]<-NA #"does not apply"

famsup<-rowMeans(cbind(famsup1, famsup2, famsup3, famsup4), na.rm = TRUE)
dat1$famsup<-famsup
```

Shared dinner with parents
*no resident parent = mean
```{r}
levels(spss1$H1WP8)
dinner<-as.numeric(substr(wave1$H1WP8, start = 2, stop = 2)) 
mean(dinner, na.rm=TRUE)
dinner[which(as.numeric(spss1$H1WP8)==10)]<-4.608819
dat1$dinner<-dinner
```

intergenerational closure
```{r}
levels(wave1$PC17)
PAclosure<-as.numeric(substr(wave1$PC17, start = 2, stop = 2))
dat1$PAclosure<-PAclosure
```

Mother/father-adolescent relationship quality
```{r}
mrel1<-as.numeric(substr(wave1$H1WP9, start = 2, stop = 2))
mrel2<-as.numeric(substr(wave1$H1WP10, start = 2, stop = 2))
mrel3<-6-as.numeric(substr(wave1$H1PF1, start = 2, stop = 2))
mrel4<-6-as.numeric(substr(wave1$H1PF4, start = 2, stop = 2))
mrel5<-6-as.numeric(substr(wave1$H1PF5, start = 2, stop = 2))
mrel<-rowMeans(cbind(mrel1, mrel2, mrel3, mrel4, mrel5), na.rm = TRUE)
mean(mrel, na.rm=TRUE)
mrel[which(as.numeric(spss1$H1WP9)==7)]<-4.430109 #no mom

drel1<-as.numeric(substr(wave1$H1WP13, start = 2, stop = 2))
drel2<-as.numeric(substr(wave1$H1WP14, start = 2, stop = 2))
drel3<-6-as.numeric(substr(wave1$H1PF23, start = 2, stop = 2))
drel4<-6-as.numeric(substr(wave1$H1PF24, start = 2, stop = 2))
drel5<-6-as.numeric(substr(wave1$H1PF25, start = 2, stop = 2))
drel<-rowMeans(cbind(drel1, drel2, drel3, drel4, drel5), na.rm = TRUE)
mean(drel, na.rm=TRUE)
drel[which(as.numeric(spss1$H1WP13)==7)]<-4.243704

dat1$mrel<-mrel
dat1$drel<-drel
```

Mother/father-adolescent shared activities (in general)
* no resident mom/dad = mean
```{r}
mtime1<-as.numeric(substr(wave1$H1WP17A, start = 2, stop = 2))
mtime2<-as.numeric(substr(wave1$H1WP17B, start = 2, stop = 2))
mtime3<-as.numeric(substr(wave1$H1WP17C, start = 2, stop = 2))
mtime4<-as.numeric(substr(wave1$H1WP17D, start = 2, stop = 2))
mtime5<-as.numeric(substr(wave1$H1WP17E, start = 2, stop = 2))
mtime6<-as.numeric(substr(wave1$H1WP17F, start = 2, stop = 2))
mtime7<-as.numeric(substr(wave1$H1WP17G, start = 2, stop = 2))
mtime8<-as.numeric(substr(wave1$H1WP17H, start = 2, stop = 2))
mtime9<-as.numeric(substr(wave1$H1WP17I, start = 2, stop = 2))
mtime10<-as.numeric(substr(wave1$H1WP17J, start = 2, stop = 2))
mtime<-mtime1+mtime2+mtime3+mtime4+mtime5+mtime6+mtime7+mtime8+mtime9+mtime10
mean(mtime, na.rm = TRUE)
mtime[which(as.numeric(spss1$H1WP17A)==4)]<- 3.951984 #no mom
summary(as.factor(mtime))

dtime1<-as.numeric(substr(wave1$H1WP18A, start = 2, stop = 2))
dtime2<-as.numeric(substr(wave1$H1WP18B, start = 2, stop = 2))
dtime3<-as.numeric(substr(wave1$H1WP18C, start = 2, stop = 2))
dtime4<-as.numeric(substr(wave1$H1WP18D, start = 2, stop = 2))
dtime5<-as.numeric(substr(wave1$H1WP18E, start = 2, stop = 2))
dtime6<-as.numeric(substr(wave1$H1WP18F, start = 2, stop = 2))
dtime7<-as.numeric(substr(wave1$H1WP18G, start = 2, stop = 2))
dtime8<-as.numeric(substr(wave1$H1WP18H, start = 2, stop = 2))
dtime9<-as.numeric(substr(wave1$H1WP18I, start = 2, stop = 2))
dtime10<-as.numeric(substr(wave1$H1WP18J, start = 2, stop = 2))
dtime<-dtime1+dtime2+dtime3+dtime4+dtime5+dtime6+dtime7+dtime8+dtime9+dtime10
mean(dtime, na.rm = TRUE)
dtime[which(as.numeric(spss1$H1WP18A)==4)]<- 2.910392 #no dad
summary(as.factor(dtime))

dat1$mactiv<-mtime
dat1$dactiv<-dtime
```

Parental control
* reverse-coded
*no resident parent = mean
```{r}
control1<-1-as.numeric(substr(wave1$H1WP1, start = 2, stop = 2))
control2<-1-as.numeric(substr(wave1$H1WP2, start = 2, stop = 2))
control3<-1-as.numeric(substr(wave1$H1WP3, start = 2, stop = 2))
control4<-1-as.numeric(substr(wave1$H1WP4, start = 2, stop = 2))
control5<-1-as.numeric(substr(wave1$H1WP5, start = 2, stop = 2))
control6<-1-as.numeric(substr(wave1$H1WP6, start = 2, stop = 2))
control7<-1-as.numeric(substr(wave1$H1WP7, start = 2, stop = 2))
control<-rowSums(cbind(control1, control2, control3, control4, control5, control6, control7), na.rm = TRUE) 
mean(control, na.rm=TRUE)
control[which(as.numeric(spss1$H1WP1)==4)]<-1.807349
summary(as.factor(control))
dat1$control<-control
```

mother/father supervision
* no mom/dad = mean
```{r}
#most/all of the time/takes = 1; other = 0
levels(wave1$H1RM11)
mspv1<-as.numeric(substr(wave1$H1RM11, start = 2, stop = 2))
mspv1[which(mspv1==1|mspv1==2|mspv1==6)]<-1
mspv1[which(mspv1==3|mspv1==4|mspv1==5)]<-0
mspv2<-as.numeric(substr(wave1$H1RM12, start = 2, stop = 2))
mspv2[which(mspv2==1|mspv2==2|mspv2==6)]<-1
mspv2[which(mspv2==3|mspv2==4|mspv2==5)]<-0
mspv3<-as.numeric(substr(wave1$H1RM13, start = 2, stop = 2))
mspv3[which(mspv3==1|mspv3==2)]<-1
mspv3[which(mspv3==3|mspv3==4|mspv3==5)]<-0
mspv<-rowSums(cbind(mspv1, mspv2, mspv3), na.rm = TRUE)
mean(mspv)
mspv[which(as.numeric(spss1$H1RM11)==8)]<-2.090406

dspv1<-as.numeric(substr(wave1$H1RF11, start = 2, stop = 2))
dspv1[which(dspv1==1|dspv1==2|dspv1==6)]<-1
dspv1[which(dspv1==3|dspv1==4|dspv1==5)]<-0
dspv2<-as.numeric(substr(wave1$H1RF12, start = 2, stop = 2))
dspv2[which(dspv2==1|dspv2==2|dspv2==6)]<-1
dspv2[which(dspv2==3|dspv2==4|dspv2==5)]<-0
dspv3<-as.numeric(substr(wave1$H1RF13, start = 2, stop = 2))
dspv3[which(dspv3==1|dspv3==2)]<-1
dspv3[which(dspv3==3|dspv3==4|dspv3==5)]<-0
dspv<-rowSums(cbind(dspv1, dspv2, dspv3), na.rm = TRUE)
mean(dspv)
dspv[which(as.numeric(spss1$H1RF11)==8)]<-1.100092

dat1$mspv<-mspv
dat1$dspv<-dspv
```
### Parental involvement with education
Mother/father involvement with schoolwork
* no resident mom/dad = mean
```{r}
minvolve1<-as.numeric(substr(wave1$H1WP17H, start = 2, stop = 2))
minvolve2<-as.numeric(substr(wave1$H1WP17I, start = 2, stop = 2))
minvolve3<-as.numeric(substr(wave1$H1WP17J, start = 2, stop = 2))
minvolve<-minvolve1+minvolve2+minvolve3
mean(minvolve, na.rm = TRUE)
minvolve[which(as.numeric(spss1$H1WP17H)==4)]<-1.280745 #no mom
summary(as.factor(minvolve))

dinvolve1<-as.numeric(substr(wave1$H1WP18H, start = 2, stop = 2))
dinvolve2<-as.numeric(substr(wave1$H1WP18I, start = 2, stop = 2))
dinvolve3<-as.numeric(substr(wave1$H1WP18J, start = 2, stop = 2))
dinvolve<-dinvolve1+dinvolve2+dinvolve3
mean(dinvolve, na.rm = TRUE)
dinvolve[which(as.numeric(spss1$H1WP18H)==4)]<- 1.068472 #no dad
summary(as.factor(dinvolve))

dat1$minvolve<-minvolve
dat1$dinvolve<-dinvolve
```

parent involvement in PTA
```{r}
levels(wave1$PA27A)
PAPTA<-as.numeric(substr(wave1$PA27A, start = 2, stop = 2))
dat1$PAPTA<-PAPTA
```

Parent in school fund-raising
```{r}
levels(spss1$PC28)
PAfund<-as.numeric(substr(wave1$PC28, start = 2, stop = 2))
dat1$PAfund<-PAfund
```

Parent met teachers
```{r}
levels(wave1$PC27)
PAteacher<-as.numeric(substr(wave1$PC27, start = 2, stop = 2))
dat1$PAteacher<-PAteacher
```

Adolescents' perceptions of mother/father educational expectations
* no resident mom/dad = mean
```{r}
mexp<-rowMeans(cbind(wave1$H1WP11, wave1$H1WP12), na.rm = TRUE)
mean(mexp, na.rm = TRUE)
mexp[which(as.numeric(spss1$H1WP11)==7)]<-4.341575 #no mom->0
dexp<-rowMeans(cbind(wave1$H1WP15, wave1$H1WP16), na.rm = TRUE)
mean(dexp, na.rm = TRUE)
dexp[which(as.numeric(spss1$H1WP15)==7)]<-4.363035 #no dad->0

summary(dexp)

dat1$mexp<-mexp
dat1$dexp<-dexp
```

### Family sociocultural characteristics
Parent nativity
*No mom/dad code as mean
```{r}
#youth report
YmomUSborn<-as.numeric(substr(wave1$H1RM2, start = 2, stop = 2))
mean(YmomUSborn, na.rm = TRUE)
YmomUSborn[which(as.numeric(spss1$H1RM2)==4)]<- 0.8798627
summary(as.factor(YmomUSborn))

YdadUSborn<-as.numeric(substr(wave1$H1RF2, start = 2, stop = 2))
mean(YdadUSborn, na.rm = TRUE)
YdadUSborn[which(as.numeric(spss1$H1RF2)==4)]<- 0.8700154
summary(as.factor(YdadUSborn))

dat1$mnativity<-YmomUSborn
dat1$dnativity<-YdadUSborn
```

Parent religiosity
* no religion--coded as lowest (Stokes, 2008)
* reverse-coded
```{r}
levels(wave1$PA23)
levels(spss1$PA23)
PArelig1<-5-as.numeric(substr(wave1$PA23, start = 2, stop = 2))
PArelig1[which(as.numeric(spss1$PA23)==6)]<-1
PArelig2<-5-as.numeric(substr(wave1$PA24, start = 2, stop = 2))
PArelig2[which(as.numeric(spss1$PA24)==6)]<-1
PArelig<-rowMeans(cbind(PArelig1, PArelig2), na.rm = TRUE)
dat1$PArelig<-PArelig
```

English as home language
```{r}
EnglishHome<-as.numeric(substr(wave1$H1GI10, start = 2, stop = 2))
EnglishHome[which(EnglishHome==2|EnglishHome==3)]<-0
dat1$EnglishHome<-EnglishHome
```

Adolescent nativity
```{r}
bornUS<-as.numeric(substr(wave1$H1GI11, start = 2, stop = 2))
for(i in 1:6504){
  if(!is.na(wave1$H1GI3[i])&wave1$H1GI3[i]==0){
    bornUS[i]<-1
  }
}
summary(bornUS)
dat1$nativity<-bornUS
```

Race/Ethnicity
```{r}
#Latino
latino<-as.numeric(substr(wave1$H1GI4, start = 2, stop = 2))
#White
white<-as.numeric(substr(wave1$H1GI6A, start = 2, stop = 2))
#African American
AA<-as.numeric(substr(wave1$H1GI6B, start = 2, stop = 2))
#Native American
Native<-as.numeric(substr(wave1$H1GI6C, start = 2, stop = 2))
#Asian
Asian<-as.numeric(substr(wave1$H1GI6D, start = 2, stop = 2))
#Other
Other<-as.numeric(substr(wave1$H1GI6E, start = 2, stop = 2))

#Using white as a reference
dat1$Latino<-latino
dat1$AA<-AA
dat1$Native<-Native
dat1$Asian<-Asian
dat1$other_race<-Other
```

### Family health resources and behaviors
parent self-rated health
* reverse-coded
```{r}
levels(wave1$PA58)
PAhealth<-6-as.numeric(substr(wave1$PA58, start = 2, stop = 2))
dat1$PAhealth<-PAhealth
```

parent smoke
```{r}
levels(wave1$PA64)
PAsmoke<-as.numeric(substr(wave1$PA64, start = 2, stop = 2))
PAsmoke[which(as.numeric(spss1$PA64)==4)]<-0 #legitimate skip-> no smokers in family
summary(as.factor(PAsmoke))
dat1$PAsmoke<-PAsmoke
```

Mother/father alcoholic (resident)
```{r}
BioMomAlc<-as.numeric(substr(wave1$PC49E_2, start = 2, stop = 2))
BioDadAlc<-as.numeric(substr(wave1$PC49E_3, start = 2, stop = 2))
malcohol<-rep(NA, 6504)
malcohol[(which(BioMom==1 & BioMomAlc==1))]<-1
malcohol[(which(BioMom==1 & BioMomAlc==0))]<-0
mean(malcohol, na.rm=TRUE)
malcohol[which(Mother==0)]<-0.0141125
summary(as.factor(malcohol))

dalcohol<-rep(NA, 6504)
dalcohol[(which(BioDad==1 & BioDadAlc==1))]<-1
dalcohol[(which(BioDad==1 & BioDadAlc==0))]<-0
mean(dalcohol, na.rm=TRUE)
dalcohol[which(Father==0)]<-0.06048632
summary(as.factor(dalcohol))

dat1$malcohol<-malcohol
dat1$dalcohol<-dalcohol
```

Mother/father disabled
*No resident mom/dad code as mean
```{r}
mdisable<-as.numeric(substr(wave1$H1RM10, start = 2, stop = 2))
mean(mdisable, na.rm = TRUE)
mdisable[which(as.numeric(spss1$H1RM10)==4)]<- 0.04313021
summary(as.factor(mdisable))

ddisable<-as.numeric(substr(wave1$H1RF10, start = 2, stop = 2))
mean(ddisable, na.rm = TRUE)
ddisable[which(as.numeric(spss1$H1RF10)==4)]<- 0.05724351
summary(as.factor(ddisable))

dat1$mdisable<-mdisable
dat1$ddisable<-ddisable
```

Mother/Father obese (resident)
```{r}
BioMomObese<-as.numeric(substr(wave1$PC49A_2, start = 2, stop = 2))
BioDadObese<-as.numeric(substr(wave1$PC49A_3, start = 2, stop = 2))
MomObese<-rep(NA, 6504)
MomObese[(which(BioMom==1 & BioMomObese==1))]<-1
MomObese[(which(BioMom==1 & BioMomObese==0))]<-0
mean(MomObese, na.rm=TRUE)
MomObese[which(Mother==0)]<-0.1901437
summary(as.factor(MomObese))

DadObese<-rep(NA, 6504)
DadObese[(which(BioDad==1 & BioDadObese==1))]<-1
DadObese[(which(BioDad==1 & BioDadObese==0))]<-0
mean(DadObese, na.rm=TRUE)
DadObese[which(Father==0)]<-0.1414504
summary(as.factor(DadObese))

dat1$mobese<-MomObese
dat1$dobese<-DadObese
```

Smoker(s) in household
```{r}
HHsmoke<-as.numeric(substr(wave1$PA63, start = 2, stop = 2))
dat1$HHsmoke<-HHsmoke
```

Illegal drugs in household
```{r}
HHdrug<-as.numeric(substr(wave1$H1TO52, start = 2, stop = 2))
dat1$HHdrug<-HHdrug
```

Family access to medical care (reverse-coded)
```{r}
fammed<-5-as.numeric(substr(wave1$PA59, start = 2, stop = 2))
summary(as.factor(fammed))
dat1$fammed<-fammed
```

Merging Wave I variables & Wave IV weights
```{r}
Wave4Weights_c<-Wave4Weights[c('AID', 'GSWGT4_2')]
dat14<-merge(dat1, Wave4Weights_c, by = 'AID')
```

Check correlations to check if there's high multicollinearity
According to Brick et al., correlations higher than .8 or .9 are problematic
```{r}
for(i in 2:55){
  for (j in (i+1):56){
    r = cor(dat14[,i], dat14[,j], use="pairwise.complete.obs")
    if (r > .50){
      print(paste(r, names(dat14)[i], names(dat14)[j]))
    }
  }
}
#None of the corrs are >.80 when using Cohen & Cohen dummy variable adjustment method!
```


## 3. Selecting and creating the outcome variables (i.e., college enrollment and college graduation/ completiong).
```{r}
dat4_re<-wave4[c("AID")]
education<-as.numeric(substr(wave4$H4ED2, start = 2, stop = 3)) #highest level of education
atschool<-as.numeric(substr(wave4$H4ED6, start = 2, stop = 2))
#dichotomize into: college entry vs. not; college completion vs. not
collen_di<-rep(NA, 5114)
collen_di[which(education<6)]<-0  #34.08%
collen_di[which(education>=6)]<-1 #65.89%
summary(as.factor(collen_di))

collcom_di<-rep(NA,5114)
collcom_di[which(education<7)]<-0  #67.36%
collcom_di[which(education>=7)]<-1 #32.61%
collcom_di[which(atschool==1 & education==6)]<-NA  #for college graduation, delete those who reported 'some college' and were still attending college
summary(as.factor(collcom_di))

dat4_re$collen_di<-collen_di
dat4_re$collcom_di<-collcom_di

dat4_re<-dat4_re[!is.na(dat4_re$collen_di),]

write.csv(dat4_re, "dat4_re.csv", row.names = FALSE)
```

##4. Handling missing data and sample weighting.
Multiple imputation?
```{r}
#first, select cases where they are not missing on dat4 outcomes
dat1_re<-dat1[dat1$AID %in% dat4_c$AID,]
write.csv(dat1_re, "dat1_re.csv", row.names = FALSE)

dat1_re<-read.csv("dat1_re.csv")
#check missing values
colmiss<-sapply(dat1_re, function(x) sum(is.na (x)))/51.13 #variable-wise
summary(colmiss) #range from 0% to 22.96% (family income)
colmiss #PAfund 57.87%, PAteacher 57.83% ->eliminate

dat1_re2<-dat1_re[,-c(19,20)]
colmiss<-sapply(dat1_re2, function(x) sum(is.na (x)))/51.13 #variable-wise
summary(colmiss) #range from 0% to 22.96% (family income)

#case-wise
dat1_re2$missing<-rowSums(is.na(dat1_re2))/53
summary(dat1_re2$missing) #range from 0 to 49%

length(dat1_re2[dat1_re2$missing>.25,]$AID) #515 eliminated for over a quarter of missingness
dat1_re3<-dat1_re2[dat1_re2$missing<=.25,] #4598 cases
dat1_re3<-dat1_re3[,-55]

sum(is.na(dat1_re3[,c(2:54)]))
4590/(4598*53) #1.88% missing in total

for(i in 2:53){
  for (j in (i+1):54){
    r = cor(dat1_re3[,i], dat1_re3[,j], use="pairwise.complete.obs")
    if (r > .50){
      print(paste(r, names(dat1_re3)[i], names(dat1_re3)[j]))
    }
  }
} #still none of the variables are correlated >.80

write.csv(dat1_re3, "dat1_re3.csv", row.names = FALSE)
```

Multiple imputation using mice
```{r}
dat1_re3_toimpute<-dat1_re3[,-1] #delete ID
cols<-c("Mother" , "Father", "TwoBioParent", "PAassistance", "welfare", "PAeconhard", "PAPTA", "mnativity", "dnativity" , "PAsmoke", "malcohol", "dalcohol", "mobese", "dobese"  ,"mdisable" , "ddisable", "HHsmoke", "HHdrug", "EnglishHome", "biosex", "Latino", "AA", "Native", "Asian", "other_race", "nativity" )
dat1_re3_toimpute[cols] <- lapply(dat1_re3_toimpute[cols], factor)
library(mice)
#rm(list=setdiff(ls(), "dat1_re3_toimpute"))
tempData_re<-mice(dat1_re3_toimpute, m = 5, maxit=500, seed = 500, 
               method = c("logreg", "logreg", "logreg", "pmm", "pmm", "pmm",
                          "pmm", "pmm", "pmm", "pmm", "pmm","logreg",
                          "logreg", "logreg", "pmm", "pmm","logreg","pmm",
                          "pmm","pmm","pmm","pmm","pmm","pmm",
                          "pmm","pmm","pmm","pmm","pmm","polyreg",
                          "polyreg","pmm","pmm","logreg","polyreg", "polyreg",
                          "polyreg","polyreg","polyreg","polyreg","pmm","logreg",
                          "logreg","pmm","pmm","","pmm","logreg",
                          "logreg","logreg","logreg","logreg","logreg"))
#'logreg' unable to predict 'EnglishHome' (only 1 NA)-> change to use 'pmm' and then later decide
saveRDS(tempData_re, "tempData_re.rds")

densityplot(tempData_re)
plot(tempData_re)

completedData_re<-complete(tempData_re, 'long')

impdat1_re<-completedData_re[completedData_re$.imp==1, -2]
impdat2_re<-completedData_re[completedData_re$.imp==2, -2]
impdat3_re<-completedData_re[completedData_re$.imp==3, -2]
impdat4_re<-completedData_re[completedData_re$.imp==4, -2]
impdat5_re<-completedData_re[completedData_re$.imp==5, -2]
```

```{r}
#For parsimony given two algorithms and nested cross-validation to run, here just take the mean or majority vote
dat1_re3_mi<-dat1_re3
factorname<-c("Mother" , "Father", "TwoBioParent", "PAassistance", "welfare", "PAeconhard", "PAPTA", "mnativity", "dnativity" , "PAsmoke", "malcohol", "dalcohol", "mobese", "dobese"  ,"mdisable" , "ddisable", "HHsmoke", "HHdrug", "EnglishHome", "biosex", "Latino", "AA", "Native", "Asian", "other_race", "nativity" )
for(j in 2:54){
  for(i in 1:4598){
    if(is.na(dat1_re3_mi[i,j])){
      print(paste(i,j))
      if(names(dat1_re3_mi)[j] %in% factorname){
        dat1_re3_mi[i,j]<-as.numeric(names(which.max(table(c(c(as.numeric(as.character(impdat1_re[i,j])), as.numeric(as.character(impdat2_re[i,j])), as.numeric(as.character(impdat3_re[i,j])), as.numeric(as.character(impdat4_re[i,j])), as.numeric(as.character(impdat5_re[i,j]))))))))
      }
      else{
        dat1_re3_mi[i,j]<-mean(c(impdat1_re[i,j], impdat2_re[i,j], impdat3_re[i,j], impdat4_re[i,j], impdat5_re[i,j]))
      }
    }
  }
}
  
sapply(dat1_re3_mi, function(x) sum(is.na (x))) #all complete 
write.csv(dat1_re3_mi, "dat1_re3_mi.csv", row.names = FALSE)
```

Output data for ML analysis
```{r}
collen<-dat4_re[!is.na(dat4_re$collen_di) & dat4_re$AID %in% dat1_re3_mi$AID, c(1,2)]
collcom_re<-dat4_re[!is.na(dat4_re$collcom_di) & dat4_re$AID %in% dat1_re3_mi$AID, c(1,3)]
dat1_collen<-dat1_re3_mi
dat1_collcom<-dat1_re3_mi[dat1_re3_mi$AID %in% collcom_re$AID,]
weights4_collen<-Wave4Weights_c[Wave4Weights_c$AID%in%dat1_collen$AID,]
weights4_collcom<-Wave4Weights_c[Wave4Weights_c$AID%in%dat1_collcom$AID,]

write.csv(collen, "collen.csv", row.names = FALSE)
write.csv(collcom_re, "collcom_re.csv", row.names = FALSE)
write.csv(dat1_collen, "dat1_collen.csv", row.names = FALSE)
write.csv(dat1_collcom, "dat1_collcom.csv", row.names = FALSE)
write.csv(weights4_collen, "weights4_collen.csv", row.names = FALSE)
write.csv(weights4_collcom, "weights4_collcom.csv", row.names = FALSE)
dat1_collen<-read.csv("dat1_collen.csv")
dat1_collcom<-read.csv("dat1_collcom.csv")
```


## 5. Preliminary analysis for descriptives (weighted).
Wave I predictors
```{r}
names(dat14)
summary(dat14$GSWGT4_2)
weighted.mean(dat14$YAge, dat14$GSWGT4_2, na.rm=TRUE)
weighted.mean(dat14$PAage, dat14$GSWGT4_2, na.rm=TRUE)
library(SDMTools)
wt.sd(dat14$YAge, dat14$GSWGT4_2)
wt.sd(dat14$PAage, dat14$GSWGT4_2)

library(weights)
wpct(dat14$biosex, weight = dat14$GSWGT4_2)
wpct(dat14$Latino, weight = dat14$GSWGT4_2)
wpct(dat14$AA, weight = dat14$GSWGT4_2)
wpct(dat14$Native, weight = dat14$GSWGT4_2)
wpct(dat14$Asian, weight = dat14$GSWGT4_2)
wpct(dat14$other_race, weight = dat14$GSWGT4_2)
Nonwhite<-dat14$Latino+dat14$AA+dat14$Native+dat14$Asian+dat14$other_race
wpct(Nonwhite, weight = dat14$GSWGT4_2)

PArelation<-as.numeric(wave1[wave1$AID %in% dat14$AID, ]$PC1)
wpct(PArelation, weight = dat14$GSWGT4_2)
8.902268e-01+2.010799e-02+1.501366e-02+1.745411e-03
```

Wave IV outcome variables
```{r}
#outcome prediction chance levels
(sum(collen$collen_di==0)/length(collen$collen_di))^2 + (sum(collen$collen_di==1)/length(collen$collen_di))^2  #54.82%
(sum(collcom_re$collcom_di==0)/length(collcom_re$collcom_di))^2 + (sum(collcom_re$collcom_di==1)/length(collcom_re$collcom_di))^2  #54.24%

#chance levels-weighted!
collen<-read.csv("collen.csv")
weights4_collen<-read.csv("weights4_collen.csv")
wpct(collen$collen_di, weight = weights4_collen$GSWGT4_2) # 0.3705646 0.6294354 
collcom_re<-read.csv("collcom_re.csv")
weights4_collcom<-read.csv("weights4_collcom.csv")
wpct(collcom_re$collcom_di, weight = weights4_collcom$GSWGT4_2)
#0.6676525 0.3323475 

0.3705646^2+0.6294354^2
0.6676525^2+0.3323475^2

#feature importance range
1-(0.3705646^2+0.6294354^2)
1-(0.6676525^2+0.3323475^2)
```






